---
title: "1. Simulate killer whale diets"
subtitle: "The power to make your own practice data"
page-layout: article
author: 
  - Amy Van Cise
  - Sarah Tanja
date: "2026-01-20"
draft: true
date-modified: today
order: 1
format: html
bibliography: "../refs/bib.bibtex"
editor: 
  markdown: 
    wrap: 72
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output))      
```

# Background

Additional reading:

-   [What's for dinner? Scientists unearth key clues to cuisine of
    resident killer
    whales](https://www.washington.edu/news/2024/09/19/killer-whale-cuisine/)
-   [Spatial and seasonal foraging patterns drive diet differences among
    north Pacific resident killer whale
    populations](https://royalsocietypublishing.org/rsos/article/11/9/rsos240445/92991/Spatial-and-seasonal-foraging-patterns-drive-diet)
    @vancise2024

### Generate KW diet data

# Load libraries

```{r}
library(tidyverse)
library(ggridges)
```

```{r}
set.seed(42)

sample.df <- data.frame(Sample = paste0("Sample", seq(1,300,1)),
                      Ind = paste0("Ind", sample(1:30, 300, TRUE)),
                      Month = as.factor(sample(1:12, 300, TRUE)))
```

```{r}
ggplot(sample.df, aes(x = Ind, fill = Month)) +
  geom_bar(stat = "count")
```

#### simulate diet

```{r}
alpha_base <- c(8, 4, 3, 0.5, 0,
  0.7, 0.5, 0.3, 0.2)
```

```{r}
seasonal_effects <- matrix(1, nrow = 9,
                           ncol = 12,
                           dimnames = list(c("Chinook salmon",
                            "chum salmon",
                            "coho salmon",
                            "steelhead salmon",
                            "pink salmon",
                            "sablefish",
                            "lingcod",
                            "arrowtooth flounder",
                            "Pacific halibut"),
                            1:12))
```

# Example seasonal structure

```{r}
seasonal_effects["Chinook salmon", c(6,7)]  <- 2.5  
seasonal_effects["Chinook salmon", c(1,11,12)]  <- 0.1
seasonal_effects["chum salmon", c(11,12)]  <- 1.3
seasonal_effects["chum salmon", c(5,6,7)]  <- 0
seasonal_effects["coho salmon", c(8,9)]   <- 2.0 
seasonal_effects["coho salmon", c(12,1,2)]   <- 0 
seasonal_effects["sablefish", c(10, 11)]  <- 7
seasonal_effects["sablefish", c(1:8)]  <- 0
seasonal_effects["lingcod", c(1,2,3)]    <- 7
seasonal_effects["lingcod", c(4:11)]    <- 0
seasonal_effects["Pacific halibut", c(2,3,4)] <- 8
seasonal_effects["Pacific halibut", c(6,7)] <- 0
seasonal_effects["arrowtooth flounder", c(6,7)] <- 0
seasonal_effects["arrowtooth flounder", c(10,11,12)] <- 10
```

# background noise

```{r}
seasonal_effects <- seasonal_effects *
  matrix(runif(9 * 12, 0.9, 1.1), 9, 12)
```

# Dirichlet distribution function

```{r}
rdirichlet_base <- function(n, alpha) {
  k <- length(alpha)
  x <- matrix(
    rgamma(n * k, shape = alpha, rate = 1),
    ncol = k,
    byrow = TRUE
  )
  x / rowSums(x)
}
```

```{r}
alpha_by_month <- lapply(1:12, function(m) {
  alpha_base * seasonal_effects[, m]
  })
```

```{r}
diet_props <- matrix(NA, nrow(sample.df), 9)
```

```{r}
for (i in seq_len(nrow(sample.df))) {
  m <- as.integer(sample.df$Month[i])
  diet_props[i, ] <- rdirichlet_base(1, alpha_by_month[[m]])
}
```

```{r}
colnames(diet_props) <- c("Chinook salmon",
                          "chum salmon",
                          "coho salmon",
                          "steelhead salmon",
                          "pink salmon",
                          "sablefish",
                          "lingcod",
                          "arrowtooth flounder",
                          "Pacific halibut")
```

```{r}
diet.df <- cbind(sample.df, diet_props) %>% 
  pivot_longer("Chinook salmon":"Pacific halibut",
               names_to = "species",
               values_to = "propDiet") %>% 
  mutate(propDiet = ifelse(propDiet < 0.01, 0, propDiet)) %>% 
  mutate(season = case_when(Month %in% c(1:3)~ "Winter",
                            Month %in% c(4:6)~ "Spring",
                            Month %in% c(7:9)~"Summer",
                            Month %in% c(10:12)~ "Autumn")) 
```


Write out diet.df
```{r}
write_csv(diet.df, "../output/KW_diet_simulated_data.csv")
```


```{r}
nSpec_samp <- diet.df %>%   
  group_by(Sample, Month, season) %>% 
  summarise(nSpecies = sum(propDiet > 0))

```

#### Some potential stats and figures ------------------------------------------

# All of these are based on number of species per month/season

```{r}
test <- aov(nSpecies ~ season, nSpec_samp)
summary(test)
```

```{r}
month.test <- aov(nSpecies ~ Month, nSpec_samp)
summary(month.test)
```

```{r}
library(dunn.test)
posthoc <- dunn.test(nSpec_samp$nSpecies, nSpec_samp$Month, list = TRUE)
```

```{r}
posthoc_results <- data.frame(Comparison = posthoc$comparisons,
  Z_Statistic = posthoc$Z,
  P_adjusted = posthoc$P.adjusted) %>% 
  filter(P_adjusted < 0.05) %>% 
  separate(Comparison, into = c("Month1", "Month2"), sep = " - ") %>% 
  pivot_longer(Month1:Month2, names_to = "ComparisonLevel", values_to = "Month") %>%
  group_by(Month) %>% 
  summarize(nSig = n())
```

```{r}
ggplot(nSpec_samp, aes(x = Month, y = nSpecies)) +
  geom_boxplot() +
  geom_point(data = posthoc_results %>% filter(nSig > 8), 
              aes(x = Month, y = 7), color = "red", shape = 8, size = 5)
```

```{r}
nSpecies.month <- diet.df %>% group_by(Month, season) %>% 
  filter(propDiet > 0) %>% 
  distinct(species) %>% 
  summarize(nSpecies = n())
```

```{r}
nSpecies.season <- diet.df %>% 
  group_by(season) %>% 
  filter(propDiet > 0) %>% 
  distinct(species) %>% 
  summarize(nSpecies = n())
```

# This is looking at a prop per species, e.g. Chinook

```{r}
propChinook <- diet.df %>% filter(species == "Chinook salmon")

chintest <- aov(propDiet ~ season, propChinook) 

summary(chintest)
```

```{r}
chin.lm <- lm(propDiet ~ as.numeric(Month), propChinook) 
summary(chin.lm)
```

```{r}
ggplot(diet.df, aes(x = as.numeric(Month), y = propDiet, color = species, fill = species)) + 
  geom_point() + 
  geom_smooth(span = 0.8, alpha = 0.5, linewidth = 0.5) + 
  facet_wrap(~species) + 
  theme_ridges() + 
  theme(legend.position = "top", text = element_text(size = 11)) +
  labs(x = "Diet proportion", y = "Month") + 
  guides(color = guide_legend(nrow = 2), 
         fill = guide_legend(nrow = 2))
```

```{r}
ggplot(diet.df %>% 
        filter(grepl("salmon", species)), 
        aes(x = propDiet, y = Month, 
            color = species, fill = species)) + 
        geom_density_ridges(scale = 1.2, alpha = 0.3, 
                          from = 0, to = 1) + 
        theme_ridges() + 
        theme(legend.position = "top", 
              text = element_text(size = 11)) + 
        labs(x = "Diet proportion", y = "Month") + 
        guides(color = guide_legend(nrow = 2), 
               fill = guide_legend(nrow = 2)) + xlim(0,1)
```

```{r}
ggplot(diet.df, 
       aes(x = Month, y = propDiet, 
           color = species, fill = species)) + 
       geom_boxplot()
```


```{r}
diet_summary <- diet.df %>% 
  group_by(Ind, species) %>% 
  summarize(meanDiet = mean(propDiet),
            sdDiet = sd(propDiet))

diet_summary
```

```{r}
ggplot(diet_summary, aes(x = meanDiet, y = species, fill = species))+
  geom_density_ridges()
```

Q1. Does southern resident killer whale diet change throughout the year? 

```{r}
ggplot(diet_summary, aes(x = species, y = mean_prop, fill = season)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = mean_prop - sd_prop, ymax = mean_prop + sd_prop),
                position = position_dodge(0.9), width = 0.25) +
  theme_minimal() +
  labs(x = "Species", y = "Mean Proportion in Diet") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Q2. Does chinook make up the majority of SRKW diet compared to other species?

```{r}
diet_chinook <- diet.df %>% 
  group_by(Ind, species) %>% 
  summarize(meanDiet = mean(propDiet),
            sdDiet = sd(propDiet)) %>%
  mutate(chinookyesno = ifelse(species == "Chinook salmon", "yes", "no"))
```


```{r}
ggplot(diet_chinook, aes(x = chinookyesno, y = meanDiet, fill = chinookyesno)) +
  geom_boxplot() +
  theme_minimal() +
  labs(x = "Chinook in Diet", y = "Mean Proportion in Diet") +
  theme(legend.position = "none")
```
anova
```{r}
chinook_anova <- aov(meanDiet ~ chinookyesno, data = diet_chinook)
summary(chinook_anova)
```
- facet_wrap
- filter

- anova or lm

- group_by
- summarise
- mutate
