---
title: "eDNA"
draft: true
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
  )
```

# Background

# Step 1. Hypotheses and variables

## Setup coding environment

- Create a new folder within your R Project for this week's analysis!
- Download the dataset `eDNA_MM_detections.csv` from the course Canvas page and save it in a `data` folder within your new analysis folder.

::: callout-important
Before starting, make sure you have the `ggOceanMaps` package installed!
You can install it by typing `install.packages("ggOceanMaps")` in the R console.
You should only need to do this once! R may ask you to restart after installation... Save your .Rmd file and restart R if prompted.
Remember.. you will need to install the package before attempting to load the package from your library...
:::

- Load libraries
```{r}
library(tidyverse)
library(ggOceanMaps) # we will need this new package to build a map!
library(raster) # for working with spatial data, dependency for marmap
library(marmap) # for bathymetry data
library(plotly) # for making an interactive 3D plot!
```
s
- Load data using `read_csv("path/to/your/file.csv")`
```{r}
detect_data <- read_csv("../data/eDNA_MM_detections.csv")
```

- Checkout your data! 
  - What are the variables? 
  - How many observations are there?
  - Use helper functions like:
    - `head(your_data)` to see the first few rows of your data
    - `glimpse(your_data)` to see the structure of your data
    - `summary(your_data)` to see summary statistics of your data
    
# Step 2. Wrangle and visualize data

```{r}
#1. Wrangle: use filter to just look at just positive detections

detect_data_positive <- detect_data %>% 
  filter(Detected == 1)
```

::: callout-tip
How many observations were filtered out when we filtered our rows (aka number of observations) to only include positive detections?
Some options for you:
- Look at the number of observations in the global environment view pane and compere the dataframes before and after filtering.
- Look at the dimensions of the data before and after filtering using the `dim(your_data)` function. 
- Or use the `nrow(your_data)` function to count the number of rows in a dataframe.
:::

```{r}
glimpse(detect_data_positive)
```


```{r, fig.width = 12}
#2a Visualize: plot MM detections by depth
ggplot(detect_data_positive, aes(y = common_name, x = depth, 
                            fill = Family, color = Family)) +
  geom_count(alpha = 0.7) +
  coord_flip(clip = "off") +
  scale_x_reverse() +
  scale_y_discrete() +
  theme_minimal()+
  theme(
    axis.text.y = element_text(size = 10),# straight labels
    axis.text.x = element_text(size = 10, 
                               angle = 90, 
                               hjust = 1) # right aligned x-axis labels
  )
```
## Make a 2D map of the sample 
```{r}
#2b Visualize: plot MM detection 2D map


#first have them make the map
base_map <- basemap(limits = c(min(detect_data$lon),
                                    max(detect_data$lon),
                                    min(detect_data$lat),
                                    max(detect_data$lat)),
                         bathy.style = "rcb", crs = 4326,
                         rotate = FALSE)
## must download data the first time this is run each day. Select 1 for yes.
```

```{r}
#installing marmap can take 30-60 seconds
#make sure R is up to date (v 4.0.x) with "version" in console
library(raster) #may need to reinstall
library(marmap)
```


```{r}
lon1 <- max(detect_data$lon) + 2
lon2 <- min(detect_data$lon) - 1

lat1 <- min(detect_data$lat) - 2
lat2 <- max(detect_data$lat) + 2
```


get the bathymetry from NOAA. Resolution in minutes, default = 4, 
minimum = 1. Smaller = higher resolution and longer times
keep=TRUE writes downloaded data into a file
lon : E to W & lat: S to N
```{r}
bathy_map <- getNOAA.bathy(lon1=lon1, lon2=lon2, lat1=lat1, lat2=lat2,
            resolution=1, keep=TRUE) 
```


```{r, fig.height=12}
#create a ggplot object appropriate to the bathy data object
base_map <- autoplot.bathy(bathy_map, geom=c('raster'),
                show.legend=FALSE) +     #turn off legend
  scale_fill_etopo() +                   #special topographic colors
  theme(axis.title = element_blank()) +  #remove the axis titles
  scale_x_continuous(breaks=seq(-126,-124, 2),  #where to place the values
                     labels=paste0(seq(126, 124, -2),'W'), 
                     expand = c(0, 0)) + 
  scale_y_continuous(breaks=seq(38,48,2),  #where to place the values
                     labels=paste0(seq(38,48,2),'N'),
                     expand = c(0, 0))
base_map
```

```{r}
base_map <- autoplot.bathy(bathy_map, geom=c('raster'),
                show.legend=FALSE) +     #turn off legend
  scale_fill_gradient2(low = "white", mid = "white", high = "gainsboro", 
                       limits   = c(min(bathy_map), 0),  # only sea depths; land becomes NA
                       na.value = "gainsboro") +     # land colour
  theme(axis.title = element_blank()) +  #remove the axis titles
  scale_x_continuous(breaks=seq(-126,-124, 2),  #where to place the values
                     labels=paste0(seq(126, 124, -2),'W'), 
                     expand = c(0, 0)) + 
  scale_y_continuous(breaks=seq(38,48,2),  #where to place the values
                     labels=paste0(seq(38,48,2),'N'),
                     expand = c(0, 0))
base_map

```


### Add points to the map
```{r}
#then add points to the map

whale_map <- base_map +
  geom_point(data = detect_data_positive, aes(x=lon, y=lat, color = Family),
             alpha = 0.5, size = 2)

whale_map
```

```{r}
#2c Visualize: plot MM detections in 3D


#using phocids (true seals) as an example here, students can do whatever they want
phocids <- detect_data_positive %>% filter(Family == "Phocidae")
```

```{r}
phocid_3D <- plot_ly(phocids,
  x = ~lon,
  y = ~lat,
  z = ~rev(depth),
  color = ~Species,
  type = "scatter3d",
  mode = "markers") %>% #this first bit up to here is all you need. The rest makes it a bit fancier
  layout(scene = list(aspectmode = "manual",
                      aspectratio = list(x = 1, y = 3, z = 0.5), #this stretches longitude axis so that it's a bit closer to reality
                      zaxis = list(autorange = "reversed"), #this reverses the depth axis so that deeper detections are at the bottom
                      xaxis = list(title = "Longitude"),
                      yaxis = list(title = "Latitude")))

```


```{r}
phocid_3D
```

fish file is 2.3GB! taking my computer (which has 50GB RAM and 6 core CPU 2 hours to download!)
```{r}
### Now let's try looking at fishes

detect_data_fish <- read_csv("../data/eDNA_MM_fish_detections.csv")
#1 Wrangle: filter for 1 MM and pivot the fish to a single column
```


# Step 3. Model the relationship