---
title: "Population structure in Hawaiian pilot whales"
subtitle: "Reviewing the scientific method, R, `tidyverse`, and markdown"
format: html
---

# The scientific method

# Review R, tidyverse, and markdown

# Population structure in Hawaiian pilot whales

## Step 1: start a script

- Set up working environment

- Load libraries
```{r}
library(tidyverse)
```

- Import data
```{r}

```

- Make exploratory visualizations

## Step 2: write your hypotheses
- Null hypothesis (H0): 
  - There is no population structure among Hawaiian short-finned pilot whales.
- Alternative hypothesis (H1a): 
  - There is significant population structure among Hawaiian short-finned pilot whales.

## Step 3: test your hypotheses

# Lab Reflection

What does this data tell you about population structure in Hawaiian
short-finned pilot whales?

1. Include your null and alternative hypotheses, and at least one plot you made today.
2. Describe their population structure in terms of the evolutionary
processes we learned about this week.
3. Be brief! <1 page please.